# Quick start {#quick-start}

Before discussing high-dimensional model selection methods, we show some examples that motivate their practical utility and that serve as a quick start for readers who are already familiar with the basics.
We focus is on sparse inference, mainly Bayesian model selection (BMS) and averaging (BMA) and L0 criteria like the AIC, BIC or EBIC.
Our examples illustrate how to obtain:

- Information criteria for many models (using MCMC model exploration)
- Posterior model probabilities
- Marginal posterior inclusion probabilities  
- BMA point estimates and posterior intervals

For these illustrations we use R package `modelSelection`, which covers several popular models.

- Generalized linear models: linear, logistic and Poisson regression. BMS, BMA and L0 criteria [@johnson:2012; @rossell:2017; @rossell:2021].

- Linear regression with non-normal residuals [@rossell:2018b], including asymmetric Normal, Laplace and asymmetric Laplace residuals.

- Accelerated Failure Time models for right-censored survival data [@rossell:2021b].

- Bayesian inference for Gaussian graphical models

- Bayesian inference for Gaussian mixture models [@fuquene:2019].

The package's C++ implementation is not optimal, but it's designed to be minimally scalable in sparse high-dimensional settings (large $p$).
On the Bayesian side, `modelSelection` is the main package implementing **non-local priors** (@johnson:2010, @johnson:2012, @rossell:2017, Section \@ref(bms-nlp)) but other popular priors are also implemented, e.g. Zellner's and Normal shrinkage priors in regression, or Gaussian spike-and-slab priors in graphical models.
The main functions for regression-type models are `modelSelection` and `bestBIC` (along with companions like `bestEBIC`, `bestAIC` and `bestIC`).
`modelSelection_eBayes` provides an empirical Bayes counterpart to `modelSelection`. 
See `modelSelectionGGM` for Gaussian graphical models and `bfnormmix` for Gaussian mixture models.
A lot of work went into coding and maintaining the package, if you use it please cite the corresponding papers.





## Linear regression

We simulate linear regression data 
$$
y_i = \sum_{j=1}^p x_{ij} \theta_j + \epsilon_i,
$$
for $p=3$ covariates and $i=1,\ldots,100$ individuals.
We set regression coefficients $\theta_1= 1$, $\theta_2= 1$, $\theta_3=0$ and $\epsilon_i \sim N(0,1)$, and the random number seed for reproducibility.
It is good practice to store the outcome and covariates into a data frame, as we do below.

```{r quickstart-simdata, message=FALSE}
library(modelSelection)
set.seed(1234)
x <- matrix(rnorm(100*3), nrow=100, ncol=3)
theta <- matrix(c(1,1,0), ncol=1)
y <- x %*% theta + rnorm(100)
df <- data.frame(y, x)
```

### L0 criteria

`bestBIC` obtains the BIC for all models. 
As usual in R we can use formulas like `y ~ X1 + X2 + X3` (provided the variables are stored in a data frame), or simply `y ~ .` to consider all the variables in the data (other than `y`) as covariates.
An intercept is added automatically, giving a total of 4 variables and $2^4=16$ possible models (the intercept can be removed by adding -1 to the formula, as usual). `bestBIC` enumerates these 16 models and finds the model with best (lowest) BIC (when there are many covariates, MCMC is used to explore the model space). 
A much faster alternative is to use a heuristic implemented in `bestBIC_fast`, based on coordinate descent and other tricks, see Section \@ref(background-l0) (for Gaussian and binary outcomes it relies on package `L0Learn` [@hazimeh:2023]). 
We set `verbose=FALSE` to avoid printing progress information and cluttering the book.
For our simulated data the selected model matches the data-generating truth, which only features the first two covariates.

```{r quickstart-bestbic}
fit.bic <- bestBIC(y ~ ., data=df, verbose=FALSE)
print(fit.bic)
```

We list the BIC for the top 5 models (index 1 corresponds to the intercept, 2 to `x[,1]` and so on). We can also use standard functions like `summary` and `coef` to view the MLE for the best model, and `predict` to obtain predictions for new data.

```{r}
fit.bic$models[1:5,]
```



```{r}
summary(fit.bic)
```


```{r}
coef(fit.bic)
```

### Bayesian model selection

A limitation of L0 criteria is that they ignore the uncertainty in the selected model, i.e. we're not completely sure that it's the correct one.
We use BMS to assess that uncertainty, in a Bayesian sense.
BMS requires setting a prior distribution on the models and on the parameters for each model.
For now, we run `modelSelection` with default priors (Beta-Binomial prior on the models, pMOM prior with default prior precision on the coefficients).


```{r quickstart-bma}
priorCoef <- momprior()
priorModel <- modelbbprior()
fit.bms <- modelSelection(y ~ ., data=df,
                     priorCoef=priorCoef, priorModel=priorModel,
                     verbose=FALSE)
```

`postProb` shows posterior model probabilities (sorted decreasingly), and `coef` gives BMA point estimates, 0.95 posterior intervals, and marginal posterior inclusion probabilities $P(\theta_j \neq 0 \mid y)$.
Below, `phi` refers to the error variance $\sigma^2$ (which is included with probability 1).
In our example, BMS selects the right covariates and assigns high posterior probability to that solution, as one would ideally wish.

```{r}
coef(fit.bms)
```

```{r}
postProb(fit.bms)[1:5,]
```

Finally, we can use `predict` to obtain point predictions and 0.95 posterior predictive intervals.

```{r}
ypred <- predict(fit.bms)
head(ypred)
```

## Logistic regression

For binary outcomes, we simply specify `family='binomial'` (and for Poisson we specify `family='poisson'`).
We first create a binary version of our simulated outcome.

```{r}
dfbin <- transform(df, ybin = (y > 0)) |>
  dplyr::select(!y) #drop variable y
```

We next use `bestBIC` and `modelSelection` as before. The selected model is still the correct one, but the  posterior probability for (wrongly) including `x[,3]` is higher than in the linear regression data. This is intuitively expected, binary outcomes carry less information than Gaussian ones, so there is more uncertainty on the chosen model.

```{r quickstart-logreg-bic}
fit2.bic <- bestBIC(ybin ~ ., family='binomial', data=dfbin, verbose=FALSE)
print(fit2.bic)
```


```{r quickstart2}
fit2.bms <- modelSelection(ybin ~ ., data=dfbin,
                     priorCoef=priorCoef, priorModel=priorModel,
                     family='binomial', verbose=FALSE)
coef(fit2.bms)
```

## Non-Linear effects via Generalized Additive Models (GAMs)

Non-linear effects can be modeled via cubic splines using the `smooth` argument (the default is 9 knots, producing 5 columns in design matrix for each non-linear covariate).
When using the smooth argument we cannot use the `~ .` notation for including all covarites, rather we must list those for which we wish to include a non-linear effect (see the example below).
The effect of each covariate is decomposed as a linear part plus a deviation from linearity (which is forced to be orthogonal to the linear term). This is useful to identify covariates for which a linear effect is sufficient, and covariates for which there are non-linearities.
`modelSelection` considers 3 possibilities for each covariate: excluding it entirely, including only the linear effect, and including both linear and non-linear terms. For further details on this decomposition, see [@rossell:2021b].

The linear effect coefficients are displayed using the original variable names, and the non-linear coefficients with an `.s` appended. Here we have 5 columns coding for the non-linear effect, labelled as `.s1` though `.s5`. 
In our example, there is strong evidence for (correctly) including the linear effect of `X1` and `X2`, excluding their non-linear effects, and exclusing `X3` entirely.


```{r quickstart-gam}
fit.gam <- modelSelection(y ~ ., data=df,
                     smooth = ~ X1 + X2 + X3,
                     priorCoef=priorCoef,
                     priorModel=priorModel, verbose=FALSE)
```


```{r}
coef(fit.gam)
```

