# Quick start {#quick-start}

Before properly discussing high-dimensional model selection methods, we show some examples to motivate their utility and to provide a quick start for readers who are already familiar with the basics.
The examples also illustrate potential pitfalls that are not always sufficiently appreciated.

We focus on sparse inference methods, mainly Bayesian model selection (BMS) and averaging (BMA), L0 criteria like the BIC or EBIC, L1 (LASSO) and adaptive LASSO.
For the illustrations we use R package modelSelection.
The main functions for regression-type models are `modelSelection` and `bestBIC` (along with companions like `bestEBIC`, `bestAIC` and `bestIC`).
`modelSelection_eBayes` provides an empirical Bayes counterpart to `modelSelection`. 
See `modelSelectionGGM` for Gaussian graphical models and `bfnormmix` for Gaussian mixture models.



## Popular methods may fail  {#quick-lasso-failure}

We consider a deceivingly simple setting with $p=3$ covariates and a moderately large sample size $n=200$, where popular high-dimensional methods fail to recover the truly active covariates.
Suppose that data are truly generated as follows:
\begin{align*}
 x_{i1} & \sim N(0,1) \\
 x_{i2} & \sim N(0,1) \\
 x_{i3} & \sim N(x_{i1} - x_{i2}, \sigma=0.2) \\
 y_i    & \sim N(x_{i1} - x_{i2}, \sigma=0.5)
\end{align*}
with independence across $i=1,\ldots,n$.
That is, $y_i$ truly depends on $x_{i1}$ and $x_{i2}$ but not on $x_{i3}$.
The complicating issue is that $x_{i3}$ is a near-copy of the mean of $y_i$, and hence the marginal correlation between $x_{i3}$ and $y_i$ is high.

```{r, message=FALSE}
library(modelSelection)
library(glmnet)
library(tidyverse)
set.seed(1234) 
n <- 200
x1 <- rnorm(n)
x2 <- rnorm(n)
x3 <- x1 - x2 + rnorm(n,sd=.2)
x <- cbind(x1, x2, x3)
y <- x1 - x2 + rnorm(n)
df <- data.frame(y, x1, x2, x3)
```

To show that this is actually an easy exercise, we fit via least-squares the model that includes all covariates. The results correctly suggest that the third covariate could be conceivably dropped, whereas covariates 1-2 are highly statistically significant.

```{r}
summary(lm(y ~ -1 + ., data=df))
```

The issue is that when the number of covariates $p$ is large, possibly larger than the sample size $n$, fitting a model that includes all parameters is less reliable (the parameter estimates are less precise).
High-dimensional methods use other strategies to identify the truly active covariates.
A popular strategy is the LASSO (or $L_1$ regularization) of [@tibshirani:1996], which can quickly obtain parameter estimates for a sequence of so-called regularization hyper-parameter values.
Let us obtain the sequence of LASSO solutions in our data. None of these models are the data-generating truth.
As one increases the LASSO penalization parameter, the path of solutions moves from the model excluding all covariates to first including $x_{i3}$ (the variable most correlated with $y_i$), and subsequently to adding $x_{i1}$ and $x_{i2}$, without ever dropping $x_{i3}$. That is, the true model is not recovered by any value of the penalization parameter.


```{r}
fit <- glmnet(x=x, y=y, family='gaussian', alpha=1, intercept=FALSE)
models <- unique(t(as.matrix(coef(fit)) != 0))
models
```
In contrast, the Bayesian information criterion (BIC) of @schwarz:1978, defined in Section \@ref(l0-basics), correctly identifies the true model for these data.
`bestBIC` enumerates all $2^3=8$ models resulting from the inclusion/exclusion of the 3 covaraites, and finds the model with best (lowest) BIC (when there are many covariates, MCMC is used to explore the model space). 
We use the R formula like `y ~ -1 + .` to consider all the variables in the data (other than `y`) as covariates,
and set `-1` to exclude the intercept, which is not needed in this example).
A much faster alternative is to use heuristics implemented in `bestBIC_fast`, based on the LASSO and adaptive LASSO solution paths, coordinate descent and other tricks, see Section \@ref(background-l0) (for Gaussian and binary outcomes it relies on package `L0Learn` [@hazimeh:2023]). 
We set `verbose=FALSE` to avoid printing progress information and cluttering the book.


```{r}
fit <- bestBIC(y ~ -1 + ., data=df, verbose=FALSE)
print(fit)
```

Fitting the selected model via least-squares shows that both truly active covariates are highly statistically significant.

```{r}
summary(fit)
```




Above we used a single simulated dataset for illustration, but this phenomenon occurs more generally.
The table below shows the proportion of times that each covariate was selected among 1,000 simulations, for $n=200$ and $n=1000$, when using the BIC, LASSO and adaptive LASSO (penalization parameter chosen by BIC).
It also shows results for the fast model search heuristic of @hazimeh:2023 (R package L0Learn). 
The code to reproduce these results can be unfolded by clicking below.



<details>
<summary>Click to show code</summary>

```{r, eval=FALSE}
#Function to simulate one dataset of size n
simdata <- function(n) {
  x1 <- rnorm(n)
  x2 <- rnorm(n)
  x3 <- x1 - x2 + rnorm(n,sd=.2)
  x <- cbind(x1, x2, x3)
  y <- x1 - x2 + rnorm(n)
  df <- data.frame(y, x1, x2, x3)
  return(df)
}

#Function to simulate nsims datasets of size n and obtain
#the proportion of times that each covariate is selected
proportionSelected <- function(nsims, n) {
  sel.L0Learn <- sel.L1 <- sel.adaL1 <- sel.bic <- matrix(NA, nrow=nsims, ncol=3)
  for (i in 1:nsims) {
    df <- simdata(n)
    fit.L0Learn <- bestBIC_fast(y ~ -1 + ., data=df, fastmethod='L0Learn', verbose=FALSE)
    fit.L1 <- bestBIC_fast(y ~ -1 + ., data=df, fastmethod='L1', verbose=FALSE)
    fit.adaL1 <- bestBIC_fast(y ~ -1 + ., data=df, fastmethod='adaptiveL1', verbose=FALSE)
    fit.bic <- bestBIC(y ~ -1 + ., data=df, verbose=FALSE)
    sel.L0Learn[i,] <- fit.L0Learn$varnames %in% fit.L0Learn$topmodel  
    sel.L1[i,] <- fit.L1$varnames %in% fit.L1$topmodel  
    sel.adaL1[i,] <- fit.adaL1$varnames %in% fit.adaL1$topmodel  
    sel.bic[i,] <- fit.bic$varnames %in% fit.bic$topmodel  
  }
  nn <- c('BIC', 'LASSO BIC', 'ALASSO BIC', 'L0Learn')
  propsel <- rbind(colMeans(sel.bic), colMeans(sel.L1), colMeans(sel.adaL1), colMeans(sel.L0Learn))
  propsel <- data.frame(Method=nn, propsel)
  return(propsel)
}

propsel.n200 <- proportionSelected(nsims=1000, n=200)
propsel.n1000 <- proportionSelected(nsims=1000, n=1000)
propsel.n200$n <- 200
propsel.n1000$n <- 1000

propsel <- rbind(propsel.n200, propsel.n1000)
write.table(propsel, "tables/lassofailure.txt", row.names=FALSE, col.names=TRUE, sep="\t")

```

</details>




```{r, echo=FALSE}
propsel <- read.table("tables/lassofailure.txt", header=TRUE, sep="\t")

tab1 <- propsel |>
  dplyr::filter(n == 200) |>
  dplyr::select(-n)

tab2 <- propsel |>
  dplyr::filter(n == 1000) |>
  dplyr::select(-n) 

combined_tab <- dplyr::full_join(tab1, tab2, by = colnames(propsel)[1])

combined_tab |>
  kableExtra::kable(
    col.names = c("Method", "X1", "X2", "X3", "X1", "X2", "X3"),
    align = "c",
    escape = FALSE
  ) |>
  # Add the grouped header row (must be named vector)
  # The names are the text you want to display, the values are the number of columns they span
  kableExtra::add_header_above(
    header = c(" " = 1, "n = 200" = 3, "n = 1,000" = 3),
    bold = TRUE
  ) |>
  kableExtra::kable_styling(full_width = FALSE)
```


BIC is the best method, including the truly active covariates 1-2 and excluding the truly inactive covariate 3 more frequently than the other methods.
LASSO-BIC and L0Learn wrongly include the third covariate in 50% of the simulations or more, and the situation does not improve as $n$ grows.
Interestingly, adaptive LASSO performs similarly to LASSO for $n=200$ and similarly to BIC for $n=1,000$.

**Remark:** LASSO-BIC uses BIC to select the best model, it's just that it only evaluates the BIC for models visited by the LASSO path, and the latter may not visit the data-generating model. For this reason, similar issues are encountered when setting the LASSO penalty with cross-validation (results not shown).

These findings are not a coincidence. There is theory for LASSO explaining that it can be expected to select a super-set of the truly active variables (the so-called screening property), but that this may contain false positives when the covariates are correlated in challenging ways (see Section \@ref(l0-theory)).
The adaptive LASSO consistently identifies the truly active covariates under less stringent conditions than LASSO, as illustrated in our findings for $n=1,000$ (with $n=200$ its performance leaves much to be desired).
In contrast, L0 criteria like the BIC have better theoretical properties in such settings.



## Model uncertainty via Bayesian model selection {#quick-bms}

A limitation of L0 criteria is that they ignore the uncertainty in the selected model, i.e. we're not completely sure that it's the correct one.
We use BMS to assess that uncertainty, in a Bayesian sense.
BMS requires setting a prior distribution on the models and on the parameters for each model.
For now, we run `modelSelection` with default priors (Beta-Binomial prior on the models, pMOM prior with default prior precision on the coefficients).


```{r quickstart-bma}
priorCoef <- momprior()
priorModel <- modelbbprior()
fit.bms <- modelSelection(y ~ ., data=df,
                     priorCoef=priorCoef, priorModel=priorModel,
                     verbose=FALSE)
```

`postProb` shows posterior model probabilities (sorted decreasingly), and `coef` gives BMA point estimates, 0.95 posterior intervals, and marginal posterior inclusion probabilities $P(\theta_j \neq 0 \mid y)$.
Below, `phi` refers to the error variance $\sigma^2$ (which is included with probability 1).
In our example, BMS selects the right covariates and assigns high posterior probability to that solution, as one would ideally wish.

```{r}
coef(fit.bms)
```

```{r}
postProb(fit.bms)[1:5,]
```

Finally, we can use `predict` to obtain point predictions and 0.95 posterior predictive intervals.

```{r}
ypred <- predict(fit.bms)
head(ypred)
```

## Logistic regression {#quick-logistic}

For binary outcomes, we simply specify `family='binomial'` (and for Poisson we specify `family='poisson'`).
We first create a binary version of our simulated outcome.

```{r}
dfbin <- transform(df, ybin = (y > 0)) |>
  dplyr::select(!y) #drop variable y
```

We next use `bestBIC` and `modelSelection` as before. The selected model is still the correct one, but the  posterior probability for (wrongly) including `x[,3]` is higher than in the linear regression data. This is intuitively expected, binary outcomes carry less information than Gaussian ones, so there is more uncertainty on the chosen model.

```{r quickstart-logreg-bic}
fit2.bic <- bestBIC(ybin ~ ., family='binomial', data=dfbin, verbose=FALSE)
print(fit2.bic)
```


```{r quickstart2}
fit2.bms <- modelSelection(ybin ~ ., data=dfbin,
                     priorCoef=priorCoef, priorModel=priorModel,
                     family='binomial', verbose=FALSE)
coef(fit2.bms)
```

## Non-linear effects via Generalized Additive Models (GAMs) {#quick-gam}

Non-linear effects can be modeled via cubic splines using the `smooth` argument (the default is 9 knots, producing 5 columns in design matrix for each non-linear covariate).
When using the smooth argument we cannot use the `~ .` notation for including all covarites, rather we must list those for which we wish to include a non-linear effect (see the example below).
The effect of each covariate is decomposed as a linear part plus a deviation from linearity (which is forced to be orthogonal to the linear term). This is useful to identify covariates for which a linear effect is sufficient, and covariates for which there are non-linearities.
`modelSelection` considers 3 possibilities for each covariate: excluding it entirely, including only the linear effect, and including both linear and non-linear terms. For further details on this decomposition, see [@rossell:2021b].

The linear effect coefficients are displayed using the original variable names, and the non-linear coefficients with an `.s` appended. Here we have 5 columns coding for the non-linear effect, labelled as `.s1` though `.s5`. 
In our example, there is strong evidence for (correctly) including the linear effect of `X1` and `X2`, excluding their non-linear effects, and exclusing `X3` entirely.


```{r quickstart-gam}
fit.gam <- modelSelection(y ~ ., data=df,
                     smooth = ~ x1 + x2 + x3,
                     priorCoef=priorCoef,
                     priorModel=priorModel, verbose=FALSE)
```


```{r}
coef(fit.gam)
```

